---
title: "Differential Gene Expression Analysis"
format: docx
editor: visual
---

```{r}
#| include: false
#Loading Packages
library(dplyr)
library(purrr)
library(readr)
library(edgeR)
library(stringr)
library(tools)
library(pathview)
library(AnnotationDbi)
library(org.Mm.eg.db)
```

### **Download the data**

We’ll start this section by downloading the data we generated as a class over the past week.

```{r}
#Download data
# scp ham14@144.175.88.28:"/home/johnson/rna-seq-johnsonra/*_gene_count_matrix.txt" data
```

### **Data Wrangling**

We should be ready to load the data and prepare it for our differential gene expression analysis.

```{r}
# List the files
files <- list.files("data", pattern = "*_gene_count_matrix.txt")

# Create a list of data frames
data <- purrr::map(file.path("data", files),
                   ~ read.delim(.x, header = TRUE, skip = 1) |>
                     dplyr::select(1, 6, 7)) |>
  purrr::reduce(dplyr::full_join, by = c("Geneid", "Length"))


# drop any bits of each name that aren't needed
names(data) <- substr(names(data), 1, 11)

# experimental groups
group <- read_csv("data/groups.csv") |>
    filter(aid %in% names(data)) # only keep accession IDs that were successfully aligned
```

### **Differential gene expression analysis with edgeR**

First we’ll start with a `DGEList` object and filter genes that are not abundant enough to be considered.

```{r}
#Create DGEList object
y <- DGEList(counts = dplyr::select(data, -Length, -Geneid),
             group = group$CXCR4,
             genes = data$Geneid,
             remove.zeros = TRUE)

#Filtering
keep <- filterByExpr(y)
y <- y[keep, , keep.lib.sizes=FALSE]

#Normalization
y <- calcNormFactors(y)
```

Next we’ll create a design matrix based on our experimental metadata and run the DEG analysis.

```{r}
# Design matrix
design <- model.matrix(~group$CXCR4)

# Estimate dispersion
y <- estimateDisp(y, design)

# Fit the model
fit <- glmQLFit(y, design)

# Conduct quasi-likelihood F-test
qlf <- glmQLFTest(fit, coef=2)

# Get the top differentially expressed genes
top_tags <- topTags(qlf)
```

### **MD Plot**

A mean-difference plot is a useful way to visualize the results of the differential expression analysis.

```{r}
plotMD(qlf, main="Pos vs Neg")
```

### **KEGG Pathway Analysis**

We will use the `kegga` function to perform KEGG pathway analysis on the differentially expressed genes.

```{r}
# Perform KEGG analysis
kegg_results <- kegga(qlf, species = "Mm")

# Display top pathways
topKEGG(kegg_results)
```

### **Pathview Visualization**

Finally, we could use `pathview` to visualize the differentially expressed genes on a KEGG pathway, typically the top pathway from our analysis.

```{r}
# Get the logFC values for all genes
logfc <- qlf$table$logFC

# Get the Entrez IDs
entrez_ids <- mapIds(org.Mm.eg.db, keys = qlf$genes$genes, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first")
names(logfc) <- entrez_ids

# Get the top pathway ID 
top_pathway_id <- "04010" #rownames(topKEGG(kegg_results))[1]

# Generate the pathview
pathview(gene.data = logfc, pathway.id = top_pathway_id, species = "mmu", kegg.native = TRUE)

```
